---
title: 'Tidyverse course: ggplot Code'
output:
  word_document:
    toc: yes
  html_notebook:
    toc: yes
    tov_float: yes
  html_document:
    df_print: paged
    toc: yes
---

Introduction
============

This document contains all of the code shown in the slides for the ggplot section of the tidyverse course.  A separate document has the answers to the exercises used in the course.

```{r}
library("tidyverse")
library("ggplot2")

msleep
```
```{r}
class(msleep)
```

Remove NA values from vore

```{r}
msleep %>% filter(!is.na(vore)) -> msleep.clean
msleep.clean
```


# Scatterplot

## Basic graph
```{r}
ggplot(
  msleep.clean, 
  aes(x=bodywt, y=sleep_total)
)+geom_point() -> scatterplot

scatterplot
```

If we don't want to save the filtered version we can pipe straight from the filter into ggplot without saving.

```{r}
msleep %>% 
  filter(!is.na(vore)) %>%
    ggplot(
      aes(x=bodywt, y=sleep_total)
  )+geom_point()
```

### with colours 
```{r}
ggplot(
  msleep.clean, 
  aes(x=bodywt, y=sleep_total, colour=vore)
)+geom_point()
```

### this also works
```{r}
ggplot(
  msleep.clean, 
  aes(x=bodywt, y=sleep_total)
)+geom_point(aes(colour=vore))
```


### with log: log axis

```{r}
ggplot(
  msleep.clean, 
  aes(x=bodywt, y=sleep_total, colour=vore)
)+geom_point() -> scatterplot

scatterplot+scale_x_log10()
```

### with log: log values 
```{r}
ggplot(
  msleep.clean, 
  aes(x=log(bodywt), y=sleep_total,colour=vore)
)+geom_point()
```


### with bigger points and axes and graph titles
```{r}
ggplot(
  msleep.clean, 
  aes(x=log(bodywt), y=sleep_total,colour=vore)
) +
  geom_point(size=4) +
  xlab("Log Body Weight") + 
  ylab("Total Hours Sleep") + 
  ggtitle("Some Sleep Data")  -> scatterplot

scatterplot
```

### change the general theme

```{r}
theme_set(theme_bw(base_size=18))

scatterplot+theme(plot.title = element_text(hjust = 0.5)) -> scatterplot
scatterplot

```


### change colour scheme and ticks on x-axis and y-axis and improve legend

This adds to the previous plot rather than recreating it.
```{r}
scatterplot +
  scale_colour_brewer(
    palette="Set1", 
    name="Trophic levels", 
    labels=c("Carnivore", "Herbivore", "Insectivore", "Omnivore")
  ) +
  scale_x_continuous(breaks=-5:10) +
  scale_y_continuous(breaks=seq(0,20, 2)) -> scatterplot
scatterplot

```


### change colours manually
```{r}
scatterplot +
  scale_color_manual(
    values=c("chocolate3", "chartreuse3", "darkorchid2","cyan3"),
    name="Trophic levels", 
    labels=c("Carnivore", "Herbivore", "Insectivore", "Omnivore")
  ) -> scatterplot

scatterplot
```

Tadah! So beautiful :-)


## Exercise 1

The file up_down_expression.txt contains an expression comparison dataset with
an extra column which classifies the rows into one of 3 groups (up, down or unchanging). 


Load: "up_down_expression.txt "
Check the structure of the file
Plot a scatterplot geom_point() with: 
up in red, 
down in blue 
unchanging in grey
Main title: "Expression data"
Colour legend: "Down, Unchanging and Up"
Axis labels: "Condition 1" and "Condition 2"


```{r}
expression <- read_tsv("up_down_expression.txt")
expression

```


```{r}
expression.scatter<-ggplot(expression, aes(Condition1, Condition2, colour=State))+
  geom_point()+
  scale_colour_manual(values=c("blue", "grey", "red"),
                      name="State", 
    labels=c("Down", "Unchanging", "Up"))+
  xlab("Condition 1") + 
  ylab("Condition 2") + 
  ggtitle("Expression data")+
  theme(plot.title = element_text(hjust = 0.5))
  
expression.scatter


```



Now, let's try another type a graph: a stripchart. It is similar to a scatterplot but the x variable is qualitative or categorical in nature.


# Stripchart

```{r}
  ggplot(
    msleep.clean, 
    aes(vore, sleep_total)
  )+geom_point()
```


### jitter, bigger points and colours
```{r}
ggplot(
  msleep.clean,
  aes(vore,sleep_total, colour=vore)
) + geom_point(size=4,position="jitter")
```

### control of the jitter 
```{r}
ggplot(
  msleep.clean, 
  aes(vore, sleep_total, colour=vore)
) +
  geom_jitter(
    width = 0.2,
    size=4,
    height = 0
  ) -> stripchart
stripchart

```


### add a line for the mean and a title for the y-axis 
```{r}
stripchart +
  stat_summary(
    fun.y="mean",
    geom='errorbar', 
    aes(ymin=..y.., ymax=..y..), 
    width=0.6, 
    size=1.5,
    colour="grey25"
  ) -> stripchart

stripchart
```

Little piece of code to calculate mean and SEm the tidyverse way.

```{r}
msleep.clean %>%
  group_by(vore) %>%
    summarise(sleep=mean(sleep_total), sem=sd(sleep_total)/sqrt(n()))
```


###  pretty changes 
```{r}
stripchart +
  ylab("Total Hours Sleep") +
  xlab("Trophic Levels") +
  ggtitle("Some Sleep Data") +
  scale_y_continuous(breaks=seq(0, 20, 2)) +
  scale_x_discrete(labels=c("Carnivore", "Herbivore", "Insectivore", "Omnivore")) +
  theme(legend.position = "none") -> stripchart

stripchart
```

### change order of levels on x axis, center the title and different colour scheme
```{r}
stripchart +
  scale_colour_brewer(palette="Dark2")+
  scale_x_discrete(
    limit=c("insecti", "herbi", "carni", "omni"),
    labels=c("Insectivore", "Herbivore", "Carnivore", "Omnivore"))+
      theme(plot.title = element_text(hjust = 0.5)
  ) -> stripchart
stripchart
```

```{r}
library("ggthemes")
stripchart+
  theme_wsj()+
  scale_colour_wsj("colors6")+
  theme(legend.position = "none")+
  theme(plot.title = element_text(hjust = 0.5))
  

```

Now, let's try some other data. DownloadFestival datacontains the hygiene scores (0-5) of 810 concert goers over three days of music festival.


### Reading a file and pre-processing it.

```{r}
read_csv("DownloadFestival.csv") -> festival.data
festival.data
```

```{r}
max(festival.data$day1)
```

# Histogram

### Histogram day 1

```{r}
ggplot(
  festival.data, 
  aes(day1)
)+geom_histogram()
```

Let's make it prettier.

```{r}
ggplot(
  festival.data, 
  aes(day1)
)+geom_histogram(binwidth=0.3)

```

Now change colour and add titles to axes.

```{r}
ggplot(
  festival.data, 
  aes(day1)
)+geom_histogram(binwidth=0.3, color="black", fill="yellow")+
  labs(x="Score", y="Counts")+
  theme(plot.title = element_text(hjust = 0.5))+
  ggtitle("Hygiene at Day 1") -> Day1Histogram
Day1Histogram

```

Now we want to plot all 3 days per gender. So we need to reshape the file. We are also going to remove the NAs.

```{r}

festival.data %>%
  gather(day,score,-ticknumb,-gender) -> festival.data.stack

festival.data.stack %>% filter(!is.na(score)) -> festival.data.stack
festival.data.stack

```

```{r}
ggplot(festival.data.stack,aes(score))+
  geom_histogram(binwidth=0.3, color="black", fill="yellow")+
  labs(x="Hygiene score", y="Counts")+
  facet_grid(gender~day) -> histogram.3days
histogram.3days

```

It is possible to modify the labels of the facets. here are some examples below.

```{r}
histogram.3days<-ggplot(festival.data.stack,aes(score))+
  geom_histogram(binwidth=0.3, color="black", fill="yellow")+
  labs(x="Hygiene score", y="Counts")+
  facet_grid(gender~day)+
  theme(strip.text.x = element_text(size = 16, colour = "purple", face="bold"),
        strip.text.y = element_text(size=12, face="bold"))
histogram.3days
```

Density plots as below.

```{r}

density.3days<-ggplot(festival.data.stack, aes(score))+
  geom_density(aes(group=day, fill=day), alpha=0.5)+
  facet_grid(~gender)
density.3days     
```


## Exercise 2:

Plot a stripchart representing all 3 days and each gender

```{r}
stripchart <-ggplot(festival.data.stack, aes(gender, score, colour=gender))+ 
	facet_grid(~day)+
	geom_point(position="jitter")+
	scale_colour_manual(values=c("darkorange", "darkorchid4"))+ 
	stat_summary(geom='errorbar',fun.y=mean, aes(ymin=..y.., ymax=..y..), 
		colour="black", width=0.8, size=1.5)+
            labs(x="Gender", y="Score")+
            theme(legend.position = "none")
stripchart
```

From a stripchart, we can add a line for the mean or any other descriptive geom as a stat summary.

```{r}
stripchart<-ggplot(festival.data.stack, aes(gender, score,colour=gender))+facet_grid(~day)+
  geom_point(position="jitter")+
  scale_colour_manual(values=c("darkorange", "darkorchid4"))+
  labs(x="Gender", y="Score")+
  theme(legend.position = "none")
  
stripchart
```

We saw how to add a mean:

```{r}
stripchart+
  stat_summary(fun.y="mean",geom="errorbar", aes(ymin=..y.., ymax=..y..), width=0.8, colour="black", size = 1.3)
  
  
```

Now let's add a boxplot.

```{r}
stripchart+
  geom_boxplot(alpha=0, colour="black")

```

We can make it prettier.

```{r}
stripchart+
  geom_boxplot(aes(gender, score, fill=gender), alpha=0.5, colour="black")+
    scale_fill_manual(values=c("darkorange", "darkorchid4"))
 
```

Speaking of making graphs prettier, we can improve on the boxplot.

# Boxplot

We can change order on the x-axis if we want:

boxplot+scale_x_discrete(limits=c("Male","Female"))

```{r}
boxplot<-ggplot(festival.data.stack, aes(gender,score))+
  geom_boxplot()+
  facet_grid(~day)

boxplot
```

```{r}
boxplot <-ggplot(festival.data.stack, aes(gender,score, fill=gender))+
	facet_grid(~day)+
	stat_boxplot(geom="errorbar", width=0.5)+	
	geom_boxplot(outlier.shape=8)+
	theme(legend.position = "none")+
	scale_fill_manual(values=c("sienna1","darkorchid3 "))+
	labs(x="Gender", y="Score")
boxplot
```


# Violinplot (beanplot)

```{r}
stripchart+
  geom_violin(alpha=0, colour="black")
```

Basic command

```{r}
violinplot<-ggplot(festival.data.stack, aes(gender,score))+geom_violin()+facet_grid(~day)
violinplot
```

Prettier: 

```{r}
violinplot<-ggplot(festival.data.stack, aes(gender,score,fill=gender))+
		facet_grid(~day)+
		geom_violin(trim = FALSE)+
		scale_fill_manual(values=c("goldenrod2","darkgrey"))+
		theme(legend.position="none")+
		stat_summary(fun.y=median, geom="point", size=2, color="black")+
		labs(x="Gender", y="Hygiene scores")
violinplot

```

```{r}
violinplot+geom_boxplot(width=0.3)
```

```{r}
violinplot+geom_jitter(width=0.1,size=1, shape=1)
```

# Barchart

First we want to calculate the means and sem and store the values in a file.

```{r}
festival.data.stack %>%
    group_by(gender,day) %>%
      summarise(mean=mean(score), sem=sd(score)/sqrt(n())) -> score.sem

score.sem
```

```{r}
barchart<-ggplot(score.sem, aes(day,mean, fill=gender))+
  geom_bar(stat="identity")
barchart
```

```{r}
barchart<-ggplot(score.sem, aes(day,mean, fill=gender))+
  geom_bar(stat="identity", position="dodge")+
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), position="dodge")
barchart
```

```{r}
barchart<-ggplot(score.sem, aes(day,mean, fill=gender))+
  geom_bar(position="dodge", stat="identity")+
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), position="dodge")
barchart
```
```{r}
barchart<-ggplot(score.sem, aes(day,mean, fill=gender))+
  geom_bar(position="dodge", colour="black",stat="identity",size=1)+
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=.5, position=position_dodge(0.9), size=1)+
  ylab("Mean scores")+ 
  ggtitle("Levels of hygiene over 3 days of concert")+
  theme(plot.title = element_text(hjust = 0.5))+
  theme(plot.title = element_text(size = 19))+
  theme(axis.title.x=element_blank())+
  scale_fill_manual(values=c("darkorange3", "darkorchid4"), name="Gender")
barchart
```

So beautiful!

# Linegraph

```{r}
linegraph<-ggplot(score.sem, aes(day, mean, group=gender))+
	geom_line()+
	geom_point()+
	geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem))

linegraph
```

```{r}
linegraph<-ggplot(score.sem, aes(day,mean, colour=gender, group=gender))+
  geom_line(size=1.5)+
  geom_point(size=4)+
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=.2, size=1.5)
linegraph
```

Change the position of the legend.

```{r}
linegraph<-ggplot(score.sem, aes(day,mean, colour=gender, group=gender))+
  geom_line(size=1.5)+
  geom_point(size=5)+
  geom_errorbar(aes(ymin=mean-sem, ymax=mean+sem), width=.2, size=1.5)+
  labs(x="", y="Mean scores")+
  scale_y_continuous(breaks=seq(0, 2, 0.2))+
  ggtitle("Levels of hygiene over 3 days of concert")+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_colour_manual(values=c("purple","darkorange3"), name="")+
  theme(legend.position = c(0.85, 0.9))+
  theme(legend.text=element_text(size=14))+
  theme(legend.background = element_rect(fill = "transparent"))

linegraph
```

## Exercise 3:

The file contains positional count data for 3 different datasets (a WT and two mutants). 

Plot a graph showing all 3 datasets on the same plot

Load: "chromosome_position_data.txt "
Check the structure of the file
Restructure the file from wide to long format gather()
Rename the column: "Genotype" and "Value"
Plot a basic line graph

```{r}
chromosome<-read_tsv("chromosome_position_data.txt")
chromosome

```

```{r}
chromosome %>%
  gather(Genotype, Value,-Position) -> chromosome.long
chromosome.long

```


```{r}
chromosome.linegraph<-ggplot(chromosome.long, aes(x=Position, y=Value, group=Genotype, colour=Genotype))+
geom_line(size=2)
chromosome.linegraph

```

## Exercise 4:

Plot a graph showing the relationship between age and weight for a typical baby
 over the first 9 months.

Load: weight_chart.txt 
Check the structure of the file
Plot a basic line graph
Plot a prettier version:
Change the size and the colour of the points
Change the thickness and the colour of the line
Change the y-axis: scale from 2 to 10 kgs
Change the x-axis: scale from 0 t 10 months
Change the labels on both the axis
Add a title to the graph


```{r}
weight<-read_tsv("weight_chart.txt")
weight

```

Basic graph:

```{r}
weight.linegraph<-ggplot(weight, aes(Age, Weight))+
  geom_line()+
  geom_point()
weight.linegraph

```

Pretty graph:

```{r}
weight.linegraph<-ggplot(weight, aes(Age, Weight))+
 geom_line(size=1, colour="lightblue2")+
 geom_point(shape=16, size=3, colour="darkorchid1")+
  scale_y_continuous(breaks=2:10, limits = c(2, 10))+
  scale_x_continuous(breaks=0:10, limits = c(0, 10))+
  labs(x="Age (months)", y="Weight (kg)")+
  ggtitle("Relation between age and weight")+
  theme(plot.title = element_text(hjust = 0.5))
weight.linegraph

```

## Exercise 5: 

The file brain_bodyweight.txt contains data for the log10 brain and bodyweight for a range of species, along with an SEM measure for each point. 

Plot these data on a scatterplot with error bars showing  the mean +/- SEM and the names of the datasets under each point. 

Load: brain_bodyweight.txt 
Check the structure of the file
Plot a basic graph. You will need:
geom_barh() for the horizontal error bars
geom_text() for the labels

Plot a prettier version.

```{r}
brain.bodyweight<- read_tsv("brain_bodyweight.txt")
brain.bodyweight

```

Basic

```{r}
brain.bodyweight.graph<-ggplot(brain.bodyweight, aes(x=Bodyweight, y=Brainweight))+
  geom_point()+
  geom_errorbar(aes(ymin=Brainweight-Brainweight.SEM, ymax=Brainweight+Brainweight.SEM))+
  geom_errorbarh(aes(xmin=Bodyweight-Bodyweight.SEM, xmax=Bodyweight+Bodyweight.SEM))+
  geom_text(aes(label=Species), hjust = 1.05, vjust = -0.6, size=2.7)
brain.bodyweight.graph

```


```{r}
brain.bodyweight.graph<-ggplot(brain.bodyweight, aes(x=Bodyweight, y=Brainweight))+
  geom_point()+
  geom_errorbar(aes(ymin=Brainweight-Brainweight.SEM, ymax=Brainweight+Brainweight.SEM), width=.1, size=1, colour="tomato3")+
  geom_errorbarh(aes(xmin=Bodyweight-Bodyweight.SEM, xmax=Bodyweight+Bodyweight.SEM), height=.1, size=1, colour="tomato3")+
  geom_point(size=2)+
  geom_text(aes(label=Species), hjust = 1.1, vjust = -0.6, size=2.7)
brain.bodyweight.graph

```

Prettier with ggrepel

```{r}
library("ggrepel")
ggplot(brain.bodyweight, aes(x=Bodyweight, y=Brainweight))+
  geom_errorbar(aes(ymin=Brainweight-Brainweight.SEM, ymax=Brainweight+Brainweight.SEM), 
width=.1, size=0.5, colour="grey28")+
  geom_errorbarh(aes(xmin=Bodyweight-Bodyweight.SEM, xmax=Bodyweight+Bodyweight.SEM), 
height=.1, size=0.5, colour="grey28")+
  geom_point(shape=21, size=3, colour="black", fill="maroon3")+
  geom_label_repel(aes(label = Species), box.padding=0.6, point.padding =0.5, 
fill="mintcream", segment.colour="grey", size=3) -> brain.bodyweight.graph
brain.bodyweight.graph

```

# Stacked bar: categorical data

```{r}
Changing<-read_csv("Changing.csv")
Changing
```


Basic

```{r}
stackedBar<-ggplot(Changing, aes(Type.of.Behaviour, Sample.Size, fill=Stage.of.Change))+
geom_bar(stat="identity")
stackedBar

```

Changing the order of the levels: factor(variable name, levels = c("", "" .))

Rotate the graph to read the x-axis labels:  coord_flip()


```{r}
Changing$Stage.of.Change <- factor(Changing$Stage.of.Change, levels = c("Maintenance","Action","Preparation","Contemplation","Precontemplation"))

stackedBar<-ggplot(Changing, aes(Type.of.Behaviour, Sample.Size, fill = Stage.of.Change))+
  geom_bar(stat="identity", colour="black")+
  coord_flip()
stackedBar
```

Prettier: 

```{r }
stackedBar<-stackedBar+
  labs(title="Stages for Each of the 12 Problem Behaviours", y="Sample Size", fill="Stages of Change")+
  theme(plot.title = element_text(hjust = 0.5, size=12, face="bold"))+
  theme(axis.title.y=element_blank())+
  scale_fill_brewer(palette = 4)+
  theme(axis.text.x = element_text(size=10), axis.text.y = element_text(size=9))+
   theme(legend.text=element_text(size=8), legend.title=element_text(size=10, face="bold"))+
  theme(axis.title.x = element_text(size=10))
stackedBar
```

With a divergent palette

```{r}
stackedBar+scale_fill_brewer(palette="RdYlGn", direction=-1)	

```

## Exercise 6

Let's plot the same data but as percentages

Plot the Changing data as percentages.

Change the format of the file into contingency xtabs()
Calculate the percentages prop.table()
Change the format into a dataframe  as.data.frame()
Check your file head()
Plot the data as before using a suitable divergent palette


```{r}
contingency.table100<-prop.table(xtabs(Sample.Size~Type.of.Behaviour+Stage.of.Change, Changing),1)*100
contingency.table100
```


```{r}
Changing.percent<-as.data.frame(contingency.table100)
Changing.percent

```

Plot the data as percentages.

```{r}
stackedBar.percent<-ggplot(Changing.percent,aes(Type.of.Behaviour, Freq, fill = Stage.of.Change))+
		geom_bar(stat="identity",colour="black")+
    coord_flip()+
		scale_fill_brewer(palette = "Spectral", direction=-1)+
		labs(title="Stages for Each of the 12 Problem Behaviours", y="Frequencies")+
    theme(axis.title.y=element_blank())+
		theme(plot.title = element_text(hjust = 0.5, size=12, face="bold"))+
    theme(axis.text.x = element_text(size=10), axis.text.y = element_text(size=9))+
    theme(legend.text=element_text(size=8), legend.title=element_text(size=10, face="bold"))+
    theme(axis.title.x = element_text(size=10))
stackedBar.percent

```


### To save a graph

Linegraph.saved<-ggsave(Linegraph, file="Line.png")
